Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-06-04 main.Rmd
added citation for KL divergence

\noindent One of the potential ways to evaluate this variation is by computing the maximum pairwise distances between the distributions of the measured variable across different levels of the cyclic granularities. <!--across different levels of $C_j$ and integrating it for all levels of $C_i$. Distances or similarity measures have been extensively studied and applied to fields like clustering, classification and anomaly detection. ---> We consider distances between probability distributions which lines up with our pivotal idea of being able to visualize probability distributions across pairs of cyclic granularities.
One of the common ways to measure distances between distributions is through Jenson-Shanon divergence (@Lin1991-fj) which is based on Kullback-Leibler divergence. The Jensen-Shanon distance between two probability distribution p1 and p2 is given by $$d = [D(p_1, r) + D(p_2, r)]/2 \quad where \quad r = (p_1 + p_2)/2$$ where,
 $$D(p_1,p_2) = \int^{\infty}_{-\infty}p_1(x)log\frac{p_1(x)}{p_2(x)}\,dx$$ is the Kullback-Leibler divergence between $p_1$ and $p_2$. Probability distributions are estimated through quantiles instead of kernel density so that there is minimal dependency on selecting kernel or bandwidth.

@@

\noindent The choice of aesthetic mapping and facets for visualizing data sets <$C_i$, $C_j$, $v$> leads to placing $A_k$'s in close proximity with $B_l$ representing a group/facet. A harmony pair ($C_i$, $C_j$) would be a more interesting display if the maximum pairwise distance between $A_k$'s $\forall k \in \{1, 2, \dots, K\}$ are higher for each group/facet. A probable solution for ranking interesting harmony pairs could be to arrange from highest to lowest average maximum pairwise distances. But maximum is not robust to the number of levels and seems to be higher for harmonies with higher levels. Thus these maximum pairwise distances need to be normalized in a way that eliminates the effect of different levels.<!--- \noindent The Fisher–Tippett–Gnedenko theorem states that the maximum of a sample of iid random variables after proper re-normalization can converge in distribution to only one of Weibull, Gumbel or Freschet distribution, independent of the underlying data or process. --> Suppose $d_{1},d_{2}\ldots ,d_{m}$ be a sequence of independent and identically-distributed pairwise distances, and $M_{m}=\max\{d_{1},\ldots ,d_{m}\}$. If a sequence of pairs of real numbers $(a_{m}, b_{m})$ exists such that each $a_{m}>0$ and $\lim _{{m\to \infty }}P\left({\frac  {M_{m}-b_{m}}{a_{m}}}\leq x\right)=F(x)$, where $F$ is a non-degenerate distribution function, then the limit distribution $F$ belongs to either the Gumbel, the Fréchet or the Weibull family. The normalizing constants $(a_{m}, b_{m})$ vary depending on the underlying distribution of the pairwise distances. Hence to normalize appropriately, it is important to assume a distribution of these distances.

@@

\noindent @Menendez1997-in and @Grosse2002-ex provide studies of the statistical properties of the Jensen-Shannon divergences and suggest that the theoretical asymptotic distribution of Jensen-Shannon divergence converges to a chi-squared distribution with degrees of freedom equal to the number of discretization used to estimate the continuous distribution. With large degrees of freedom, chi-squared is asymptotically normal by the CLT. Hence, the norming constants for the distribution of maximum distance are chosen to be $b_m = \Phi^{-1}(1- 1/m)$ and $a_m = 1/m\phi(b_m)$, where $\Phi^{-1}$ and $\phi$ are the quantile and density function of a standard normal distribution.

@@

4. Compute maximum within the $m$ computed distances.

5. Compute normalized maximum distance ($NM$) using appropriate norming constants so that the distribution of the normalized maximum converges to a Gumbel distribution as $m\rightarrow\infty$.

@@

7. Compute MMPD = median $(NM_1, NM_2, \dots, NM_K)$/log($K$). Median normalized distance grows at the rate of log($K$) as $K\rightarrow\infty$ and hence is normalized for the value of $K$ in MMPD.

@@

The method is not only restricted to temporal data, and can be generalized to many hierarchical granularities (with  continuous and uni-directional nature). We illustrate this with an application to the sport cricket. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In the Twenty20 format, an over will consist of 6 balls (with some exceptions), an inning is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus, similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. The idea of cyclic granularities can be likewise mapped to this hierarchy. Example granularites then include ball of the over, over of the inning and ball of the inning. Although most of these cyclic granularities are circular in design of the hierarchy, in application of the rules some granularities will are aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-over will be circular in most cases and aperiodic in others.

@@

First, it would be interesting to see if the distribution of total runs vary depending on if a team bats in the first or second inning. The Mumbai Indians (MI) and Chennai Super kings (CSK) <!--are considered one of the best teams in IPL with multiple winning titles and --> appeared in final playoffs from 2010 to 2015. We take their example in order to dive deeper into this question. <!--Circular granularities "over-of-inning" and "inning-of-match" can be computed using \ref{sec:circular-gran-def} with over as index of the tsibble.--> From Figure \ref{fig:cricex}(a), it can be observed that for the team batting in the first inning there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the teams batting in the second inning. This seem to indicate that players feel mounting pressure to score more runs as they approach towards the end of the first inning. Whereas teams batting in the second inning have a set target in mind and are not subjected to such mounting pressure and may adopt a more conservative strategy, to score runs. Thus winning teams like CSK and MI seem to employ different inning strategies when it comes to their batting order.

@@

Wickets per over are considered as aperiodic cyclic granularities with wickets as aperiodic linear granularities. These granularities do not appear in the hierarchy table since it is difficult to position them in a hierarchy. These are similar to holidays or special events in temporal data.
<!-- CONTENT CUT -->
<!--While any special event that corresponds to a time domain can be treated as an aperiodic linear granularity in a temporal case, dot balls or wickets that corresponds to certain balls (index) could be treated as aperiodic events in cricket.-->

>>>>

\noindent One of the potential ways to evaluate this variation is by computing the maximum pairwise distances between the distributions of the measured variable across different levels of the cyclic granularities. <!--across different levels of $C_j$ and integrating it for all levels of $C_i$. Distances or similarity measures have been extensively studied and applied to fields like clustering, classification and anomaly detection. ---> We consider distances between probability distributions which lines up with our pivotal idea of being able to visualize probability distributions across pairs of cyclic granularities.
One of the common ways to measure distances between distributions is through Jenson-Shanon divergence (@Lin1991-fj) which is based on Kullback-Leibler divergence (@Kullback1951-jy). The Jensen-Shanon distance between two probability distribution $p_1$ and $p_2$ is given by $$d = [D(p_1, r) + D(p_2, r)]/2 \quad where \quad r = (p_1 + p_2)/2$$ where,
 $$D(p_1,p_2) = \int^{\infty}_{-\infty}p_1(x)log\frac{p_1(x)}{p_2(x)}\,dx$$ is the Kullback-Leibler divergence between $p_1$ and $p_2$. Probability distributions are estimated through quantiles instead of kernel density so that there is minimal dependency on selecting kernel or bandwidth.

@@

\noindent The choice of aesthetic mapping and facets for visualizing data sets <$C_i$, $C_j$, $v$> leads to placing $A_k$'s in close proximity with $B_l$'s representing a group/facet. A harmony pair ($C_i$, $C_j$) would be a more interesting display if the maximum pairwise distance between $A_k$'s $\forall k \in \{1, 2, \dots, K\}$ are higher for each group/facet. A probable solution for ranking interesting harmony pairs could be to arrange from highest to lowest average maximum pairwise distances. But maximum is not robust to the number of levels and seems to be higher for harmonies with higher number of levels. Thus these maximum pairwise distances need to be normalized in a way that eliminates the effect of different levels.<!--- \noindent The Fisher–Tippett–Gnedenko theorem states that the maximum of a sample of iid random variables after proper re-normalization can converge in distribution to only one of Weibull, Gumbel or Freschet distribution, independent of the underlying data or process. --> Suppose $d_{1},d_{2}\ldots ,d_{m}$ be a sequence of independent and identically-distributed pairwise distances and $M_{m}=\max\{d_{1},\ldots ,d_{m}\}$. Then Fisher–Tippett–Gnedenko theorem [@De_Haan2007-yx] suggests that if a sequence of pairs of real numbers $(a_{m}, b_{m})$ exists such that each $a_{m}>0$ and $\lim _{{m\to \infty }}P\left({\frac  {M_{m}-b_{m}}{a_{m}}}\leq x\right)=F(x)$, where $F$ is a non-degenerate distribution function, then the limit distribution $F$ belongs to either the Gumbel, Fréchet or Weibull family. The normalizing constants $(a_{m}, b_{m})$ vary depending on the underlying distribution of the pairwise distances. Hence to normalize appropriately, it is important to assume a distribution of these distances.

@@

\noindent @Menendez1997-in and @Grosse2002-ex provide studies of the statistical properties of the Jensen-Shannon divergences and suggest that the theoretical asymptotic distribution of Jensen-Shannon divergence converges to a chi-squared distribution with degrees of freedom equal to the number of discretization used to estimate the continuous distribution. With large degrees of freedom, chi-squared is asymptotically normal by the Central Limit Theorem. Hence, the norming constants for the distribution of $M_m$ are chosen to be $b_m = \Phi^{-1}(1- 1/m)$ and $a_m = 1/m\phi(b_m)$, where $\Phi^{-1}$ and $\phi$ are the quantile and density function of a standard normal distribution. Typically normal distributions will have Gumbel limiting distribution for the maxima, which have medians that grow as $m\rightarrow\infty$ at the rate of log $m$.

<!-- Median normalized distance grows at the rate of log($K$) as $K\rightarrow\infty$ and hence is normalized for the value of $K$ in MMPD.  -->

@@

4. Identify maximum within the $m$ computed distances.

5. Compute normalized maximum distance ($NM$) using appropriate norming constants.
<!--so that the distribution of the normalized maximum converges to a Gumbel distribution as $m\rightarrow\infty$.-->

@@

7. Compute MMPD = median $(NM_1, NM_2, \dots, NM_K)$/log($K$).

@@

The method is not only restricted to temporal data, and can be generalized to many hierarchical granularities (with  continuous and uni-directional nature). We illustrate this with an application to the sport cricket. Although there is no conventional time component in cricket, each ball can be thought to represent an ordering from past to future with the game progressing forward with each ball. In the Twenty20 format, an over will consist of 6 balls (with some exceptions), an inning is restricted to a maximum of 20 overs, a match will consist of 2 innings and a season consists of several matches. Thus, similar to time, there is a hierarchy where ball is nested within overs, overs nested within innings and innings within matches. The idea of cyclic granularities can be likewise mapped to this hierarchy. Example granularites then include ball of the over, over of the inning and ball of the inning. Although most of these cyclic granularities are circular in design of the hierarchy, in application of the rules some granularities are aperiodic. For example, in most cases an over will consist of 6 balls with some exceptions like wide balls or when an inning finishes before the over finishes. Thus, the cyclic granularity ball-of-over will be circular in most cases and aperiodic in others.

@@

First, it would be interesting to see if the distribution of total runs vary depending on if a team bats in the first or second innings. The Mumbai Indians (MI) and Chennai Super kings (CSK) <!--are considered one of the best teams in IPL with multiple winning titles and --> appeared in final playoffs from 2010 to 2015. We take their example in order to dive deeper into this question. <!--Circular granularities "over-of-inning" and "inning-of-match" can be computed using \ref{sec:circular-gran-def} with over as index of the tsibble.--> From Figure \ref{fig:cricex}(a), it can be observed that for the team batting in the first inning there is an upward trend of runs per over, while there is no clear upward trend in median and quartile deviation of runs for the teams batting in the second inning. This seem to indicate that players feel mounting pressure to score more runs as they approach towards the end of the first inning. Whereas teams batting in the second inning have a set target in mind and are not subjected to such mounting pressure and may adopt a more conservative strategy, to score runs. Thus winning teams like CSK and MI seem to employ different inning strategies when it comes to their batting order.

@@

Wickets per over are considered as an aperiodic cyclic granularity with wickets as an aperiodic linear granularity. These granularities do not appear in the hierarchy table since it is difficult to position them in a hierarchy. These are similar to holidays or special events in temporal data.
<!-- CONTENT CUT -->
<!--While any special event that corresponds to a time domain can be treated as an aperiodic linear granularity in a temporal case, dot balls or wickets that corresponds to certain balls (index) could be treated as aperiodic events in cricket.-->
